---
title: "Idiographic Networks of Procrastination"
subtitle: "graphical VAR, mlVAR, and GIMME on ESM data — the `idionet` package"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    df_print: paged
    fig_width: 8
    fig_height: 6
---

```{r setup, include=FALSE}
# Global options: keep package-load chatter quiet, but DO NOT globally hide
# warnings — model-fit diagnostics (singular lmer fits, lavaan convergence) are
# suppressed locally, per chunk, only where they are expected and explained.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = TRUE,
                      fig.align = "center", dpi = 110)
library(gimme)      # plain attach: gimme registers S3 methods on load
library(idionet)    # the package this tutorial documents
library(cograph)    # renderer
```

The three methods in `idionet` answer the same question from three vantage
points: **how do a person's states drive one another over time?** They differ in
*whose* dynamics they model.

- **graphical VAR** models **one person's** time series — a fully idiographic
  network.
- **mlVAR** pools **many people** in a multilevel model, separating *within*-person
  dynamics from *between*-person differences.
- **GIMME** builds a **person-specific** directed model for each individual while
  keeping the paths shared by enough people as a group structure.

We use them on experience-sampling (ESM) data to study the affective and
self-regulatory dynamics around **procrastination**.

# The data

The bundled demonstration dataset (`inst/extdata/esm_demo.tsv`, shipped with the
package) is a **synthetic** ESM study: students sampled roughly four times a day,
each rating six 1–5 items at every beep — three self-regulation/engagement states
and three negative-affect states that, in the procrastination literature, form
the self-control–emotion cycle:

- self-regulation / engagement: `goal_directed`, `purposeful`, `attentive`
- approach affect: `excited`
- the procrastination affect: `worries`, `guilty`

```{r data}
raw <- read.delim(
  system.file("extdata", "esm_demo.tsv", package = "idionet"),
  stringsAsFactors = FALSE
)
vars <- c("goal_directed", "purposeful", "attentive",
          "excited", "worries", "guilty")

# Everything the narrative quotes is computed here, never hard-coded.
counts      <- table(raw$id)
n_students  <- length(counts)
subject_id  <- as.integer(names(which.max(counts)))   # most densely sampled
subject_n   <- as.integer(max(counts))                # its number of beeps
ml_min      <- 40L                                     # mlVAR inclusion threshold
gi_min      <- 48L                                     # GIMME inclusion threshold
n_ml        <- sum(counts >= ml_min)
n_gi        <- sum(counts >= gi_min)

cat(sprintf("%d students, %d total observations\n", n_students, nrow(raw)))
```

Each method takes the raw long data frame directly and selects which subjects to
analyse through its own arguments — `subject` (name the exact person/people) or
`min_obs` (keep everyone with at least that many observations). Counts are read
from the data frame, so no sorting or filtering happens in the analysis code.
`day` and `beep` matter: lagged effects are computed *within a day*, never across
the overnight gap, so the lag-1 relationships are real consecutive measurements.

# One person — graphical VAR

**Question:** within the most densely sampled student, which states predict
which others at the next beep, and which move together at the same beep?

We fit a graphical VAR for student **`r subject_id`**, the most densely sampled
person in the study (`r subject_n` beeps). `gamma` controls EBIC sparsity; we use
`0` (BIC) because a single person's series is short and we want to see the
structure rather than prune it away.

```{r gvar, warning=FALSE}
gv <- graphical_var(raw, vars = vars, id = "id", day = "day", beep = "beep",
                    subject = subject_id,      # study this specific student
                    n_lambda = 25, gamma = 0)
gv
```

`graphical_var()` returns two networks. The **temporal** network holds the lag-1
directed effects (including each variable's autoregression — its carry-over from
one beep to the next); the **contemporaneous** network holds the partial
correlations among the same-beep residuals.

```{r gvar-plot, fig.show="hold", out.width="49%", fig.width=6, fig.height=6}
nets <- as_netobject(gv)          # two plottable netobjects, no matrix wrangling
splot(nets$temporal)
splot(nets$contemporaneous)
```

Left: temporal (an edge runs from a state at one beep to a state at the next;
self-loops are carry-over). Right: contemporaneous (undirected same-beep partial
correlations).

```{r gvar-edges}
knitr::kable(extract_edges(nets$contemporaneous, sort_by = "weight"),
             digits = 3, caption = "Strongest same-beep partial correlations")
```

**Read it:** the same-beep network captures the moment-to-moment co-movement —
when `worries` and `guilty` rise together while the self-regulation states
(`goal_directed`, `purposeful`, `attentive`) move together, that is the
signature of procrastination affect. The temporal side is dominated by
carry-over (autoregression): states persist from beep to beep more than they
cross-predict, which is typical of short single-person series.

# The group — mlVAR

**Question:** across many students, what are the *average within-person*
dynamics, and separately, how do students who are high on one state tend to
differ on the others?

We keep students with at least `r ml_min` observations
(`r n_ml` students) and fit a multilevel VAR. Unlike the single-person model,
mlVAR returns a **third** network — the between-person level — which a single
time series cannot produce.

```{r mlvar, warning=FALSE}
fit <- build_mlvar(raw, vars = vars, id = "id", day = "day", beep = "beep",
                   min_obs = ml_min,          # students with >= ml_min beeps
                   standardize = TRUE)
fit
```

```{r mlvar-plot, fig.show="hold", out.width="32%", fig.width=5, fig.height=5}
mnets <- as_netobject(fit)        # plotting-oriented networks (predictor -> outcome)
splot(mnets$temporal)
splot(mnets$contemporaneous)
splot(mnets$between)
```

Left to right: **temporal** (average within-person lag-1 effects),
**contemporaneous** (within-person same-beep partial correlations), **between**
(how the person-level means co-vary across students).

```{r mlvar-temporal-edges}
knitr::kable(extract_edges(mnets$temporal, sort_by = "weight"),
             digits = 3, caption = "Average within-person lag-1 effects (predictor -> outcome)")
```

**Read it:** the temporal network now has cross-lagged structure that the single
person lacked — pooling `r n_ml` students estimates each lagged coefficient from
far more data. Look for self-regulation states at one beep predicting lower
`worries`/`guilty` at the next (or the reverse): that directional arrow is the
group-average procrastination dynamic. The **between** network answers a
different question entirely — students who are generally more `goal_directed`
also tend to be generally less `guilty` — a trait-level pattern, not a process.

# Person-specific structure — GIMME

**Question:** does a single model fit everyone, or does each student have their
own procrastination network — and what, if anything, is shared by the group?

GIMME searches, per person, for the lagged and contemporaneous directed paths
that improve fit, promoting a path to **group-level** when enough individuals
need it. It needs longer series, so we restrict to the densely sampled
students — those with at least `r gi_min` observations (`r n_gi` students).

```{r gimme, warning=FALSE}
gmod <- build_gimme(raw, vars = vars, id = "id", day = "day", beep = "beep",
                    min_obs = gi_min,          # students with >= gi_min observations
                    ar = TRUE, seed = 42)
gmod
```

```{r gimme-plot, fig.width=7.5, fig.height=7}
plot_gimme(gmod)        # the gimme-style network, rendered by cograph
```

`plot_gimme()` draws the GIMME network exactly as the `gimme` package does — a
single `p`-node graph where **dashed edges are lag-1 (temporal)** and **solid
edges are lag-0 (contemporaneous)**, the autoregression is a dashed self-loop,
**edge width is the proportion of students that have the path**, and **black
edges are group-level** paths (present for enough students) versus grey
individual-level ones. So the picture encodes *how prevalent* each path is across
people — not an averaged coefficient.

```{r gimme-pathcounts}
knitr::kable(as.data.frame.matrix(gmod$path_counts),
             caption = "How many students have each path (rows=outcome, cols=predictor; *lag = temporal)")
```

The plot is just this counts table made visual (width = count / number of
students). Each student's own standardized estimates live in `gmod$coefs` (one
`outcome x [lagged | contemporaneous]` matrix per person); the matrix view
`as_netobject(gmod)` returns the same prevalence as `$temporal` / `$contemporaneous`.

**Read it:** the diagonal of the counts table (the `*lag` self-paths) is the
autoregression every student shares — states carry over for everyone. Off-diagonal
paths that appear in only one or two students are genuinely **idiographic**: that
student's procrastination runs through a route the others don't have. A path
present in most students is what GIMME elevates to the group model — a candidate
*nomothetic* mechanism inside an idiographic method.

# When to use which

| You have… | You want to know… | Use |
|---|---|---|
| one person's series | that person's own dynamics | **graphical VAR** |
| many people, repeated measures | the average within-person process **and** between-person differences | **mlVAR** |
| many people, longer series | a separate network per person, plus what they share | **GIMME** |

- **graphical VAR** is the right tool when the unit of interest *is* the
  individual — clinical single-case work, n-of-1 designs, or one richly sampled
  person.
- **mlVAR** is the workhorse for ESM samples: it borrows strength across people
  (so the lagged effects are stable) and is the only one here that separates the
  within- from the between-person level.
- **GIMME** is for when you suspect people genuinely differ in *structure*, not
  just in degree — it refuses to assume one network fits all, while still
  surfacing the shared paths.

For procrastination specifically: mlVAR tells you the average emotion–regulation
cycle and how chronic procrastinators differ as people; graphical VAR and GIMME
tell you whether a *particular* student's cycle matches that average or runs its
own course — which is exactly the question that matters for an individual
intervention.

```{r sessioninfo}
sessionInfo()
```
