---
title: "idionet `build_gimme()` vs the `gimme` package"
subtitle: "Numerical equivalence and a faithful side-by-side, on the bundled ESM data"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", dpi = 110)
library(idionet)
library(cograph)

# Tiny helpers so every bracket index / order() stays inside a function body,
# never on the vignette's public surface (the tidy-code convention). Aligning
# two packages' raw matrices is the one place indexing is unavoidable.
order_panel <- function(df, by) df[do.call(order, df[by]), , drop = FALSE]
align_to    <- function(m, ref) m[rownames(ref), colnames(ref)]
```

This compares `idionet::build_gimme()` with the `gimme` package on the **bundled
demonstration dataset** (`inst/extdata/esm_demo.tsv`), and shows the corrected
plot: idionet now renders GIMME the way gimme does — a single `p`-node network
with **dashed = lag-1, solid = lag-0, width = proportion of subjects, black =
group / grey = individual** — drawn by `cograph`.

> **One alignment detail.** `gimme` lags *consecutive rows* and has no notion of
> day/session boundaries. idionet can lag *within day* (so a lag never crosses
> the overnight gap), which is more correct for ESM — but to compare the two
> engines like-for-like we turn that off here and let both lag consecutively
> (idionet via a continuous `time` index, no `day`).

# The data

```{r data}
raw <- read.delim(
  system.file("extdata", "esm_demo.tsv", package = "idionet"),
  stringsAsFactors = FALSE
)
vars <- c("goal_directed", "purposeful", "attentive",
          "excited", "worries", "guilty")

# Use the densely sampled students so gimme has enough time points per person.
counts  <- table(raw$id)
keep_id <- as.integer(names(which(counts >= 48)))
sub     <- subset(raw, id %in% keep_id)

# Order within each student and build a continuous within-person time index so
# idionet and gimme form identical lag pairs.
sub     <- order_panel(sub, c("id", "day", "beep"))
sub$.t  <- stats::ave(seq_len(nrow(sub)), sub$id, FUN = seq_along)

cat(sprintf("%d students (>= 48 obs), %d observations\n",
            length(keep_id), nrow(sub)))
```

# Run both engines

`gimme::gimme()` only runs in a top-level R session (it fails inside a `knitr`
chunk), so we run it in a **subprocess** that reads one CSV per student and
writes back its path counts, per-person betas, and a PNG of its native plot.
`idionet::build_gimme()` runs in the chunk.

```{r run, results="hide", fig.show="hide"}
indir <- file.path(tempdir(), "esm_csv"); dir.create(indir, showWarnings = FALSE)
invisible(lapply(keep_id, function(s) {
  write.csv(subset(sub, id == s, select = vars),
            file.path(indir, paste0("s", s, ".csv")), row.names = FALSE)
}))

gout     <- file.path(tempdir(), "esm_gimme_out"); dir.create(gout, showWarnings = FALSE)
art_rds  <- file.path(tempdir(), "esm_gimme_art.rds")
plot_png <- file.path(tempdir(), "esm_gimme_native.png")
runner   <- file.path(tempdir(), "run_gimme_esm.R")
writeLines(c(
  'suppressMessages(suppressWarnings(library(gimme)))',
  'a <- commandArgs(trailingOnly = TRUE)  # indir, gout, art_rds, plot_png',
  'g <- gimme(data = a[1], sep = ",", header = TRUE, out = a[2],',
  '           ar = TRUE, plot = TRUE, groupcutoff = 0.75)',
  'saveRDS(list(path_counts = as.matrix(g$path_counts),',
  '             path_est_mats = g$path_est_mats), a[3])',
  'png(a[4], width = 640, height = 640, res = 100)',
  'plot(g$group_plot_paths); dev.off()'
), runner)
status <- system2("Rscript", c(shQuote(runner), shQuote(indir), shQuote(gout),
                               shQuote(art_rds), shQuote(plot_png)),
                  stdout = FALSE, stderr = FALSE)
stopifnot(status == 0L)
gref <- readRDS(art_rds)

# idionet: continuous time index, no `day`, so lags match gimme exactly.
ido <- build_gimme(sub, vars = vars, id = "id", time = ".t",
                   ar = TRUE, groupcutoff = 0.75, seed = 1)
```

# Numerical equivalence

```{r counts}
gc <- gref$path_counts
ic <- align_to(ido$path_counts, gc)
knitr::kable(gc, caption = "gimme: path counts")
knitr::kable(ic, caption = "idionet: path counts")
# Compare values (gimme stores counts as double, idionet as integer, so a strict
# identical() would trip on type alone — every cell value is what matters).
cat("All path-count cells equal: ", all(gc == ic),
    " | max cell difference: ", max(abs(gc - ic)), "\n")
```

```{r percoef}
common <- intersect(names(gref$path_est_mats), paste0("s", keep_id))
max_abs_diff <- vapply(common, function(nm) {
  G <- as.matrix(gref$path_est_mats[[nm]])
  sid <- sub("^s", "", nm)
  I <- align_to(ido$coefs[[sid]], G)
  max(abs(G - I))
}, numeric(1))
knitr::kable(data.frame(subject = common,
                        max_abs_diff = signif(max_abs_diff, 3)),
             caption = "Max |idionet - gimme| per-subject standardized coefficient")
cat("Overall max |difference|: ", signif(max(max_abs_diff), 3), "\n")
```

# The plots, side by side

## gimme's native plot

```{r plot-gimme, out.width="60%"}
knitr::include_graphics(plot_png)
```

## idionet, rendered by cograph (the corrected, faithful view)

Same conventions as gimme: dashed = lag-1, solid = lag-0, width = proportion of
students with the path, black = group / grey = individual, autoregression as a
dashed self-loop.

```{r plot-ido, fig.width=7.5, fig.height=7}
plot_gimme(ido)
```

```{r mixed-edges}
# The exact numbers behind the plot, via the tidy edges() verb (one row per
# path, sorted by |weight|; network = temporal/contemporaneous, level =
# group/individual).
knitr::kable(edges(ido), digits = 3, row.names = FALSE,
             caption = "GIMME paths: network, proportion of students (weight), level")
```

# Takeaway

- **Numerically**, `idionet::build_gimme()` reproduces the `gimme` package on the
  bundled data: identical group path counts and per-person standardized
  coefficients (max difference above).
- **Visually**, idionet's GIMME plot now matches gimme's convention — one
  `p`-node network, dashed lag / solid contemporaneous, width = prevalence,
  group/individual colouring — rendered entirely by `cograph` via
  [plot_gimme()]. The earlier `2p`-node, averaged-coefficient picture is gone.
