---
title: "10. Comparing methods"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{10. Comparing methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 5.5
)
```

```{r libraries}
library(idiographic)
data(srl)
vars <- c("efficacy", "value", "planning", "monitoring", "effort")
```

`compare_idiographic()` runs several estimators on the same data and stacks
their summaries into **one comparison table**, so you read estimators against
each other instead of assembling the comparison by hand.

Each estimator's own arguments are passed through `estimator_args`. Here we
compare the two single-person temporal estimators on `Grace`.

```{r compare}
cmp <- compare_idiographic(
  srl, vars = vars, id = "name",
  estimators = c("var", "graphical_var"),
  estimator_args = list(
    var = list(subject = "Grace", scale = TRUE),
    graphical_var = list(subject = "Grace", n_lambda = 8, gamma = 0)
  )
)

cmp
```

The comparison itself is a tidy `data.frame`:

```{r compare-table}
as.data.frame(cmp)
```

Add more estimators by extending `estimators` and `estimator_args`. The group
methods (`mlvar`, `usem`, `gimme`) are accepted too — they are heavier, so the
full template below is shown but not run here:

```{r compare-full, eval=FALSE}
students <- subset(srl, name %in% c("Grace", "Eve", "Aisha", "Alice",
                                    "Bob", "Diana", "Frank", "Heidi"))

compare_idiographic(
  students, vars = vars, id = "name",
  estimators = c("var", "graphical_var", "mlvar", "usem", "gimme"),
  estimator_args = list(
    var = list(subject = "Grace", scale = TRUE),
    graphical_var = list(subject = "Grace", n_lambda = 8, gamma = 0),
    mlvar = list(standardize = TRUE),
    usem = list(time = "day", temporal = "ar", contemporaneous = "all",
                residual_cov = TRUE, trim = TRUE, seed = 1),
    gimme = list(time = "day", ar = TRUE, seed = 1)
  )
)
```
