---
title: "Within-Person Correlation with correlate_persons()"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Within-Person Correlation with correlate_persons()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE, fig.width = 7.2, fig.height = 4.6, dpi = 144, fig.align = "center", out.width = "100%")
source("figure-style.R")
library(idiographic)
data(srl)
analysis_data <- subset(srl, name %in% c("Aisha", "Alice", "Anika", "Astrid", "Bjorn", "Bob", "Charlie", "Diana", "Erik", "Eve", "Fatima", "Frank"))
```

## Analytic question

`correlate_persons()` estimates how two variables covary within each person.
The function takes repeated measurements, a person identifier, and at least
two variables. It returns one row per person and variable pair with the
correlation, complete-pair count, confidence interval, and p-value.

`correlate_persons()` does not pool occasions across learners. A positive
coefficient for one learner means that the learner tends to report higher
values on both variables at the same occasions. It does not mean that learners
with higher average values on one variable also have higher averages on the
other variable.

## Estimate the correlations

`correlate_persons()` computes confidence intervals with Fisher's
transformation. The `time` argument fixes row order for consistency with other
person-level analyses, although contemporaneous correlation itself is
unchanged by row order. Table 1 keeps two learners and three variable pairs.

```{r correlation-table}
correlations <- correlate_persons(analysis_data, id = "name", vars = c("efficacy", "monitoring", "effort"), time = "day", subject = c("Aisha", "Alice"))
correlations
```

`correlate_persons()` finds positive associations among all three variables
for Aisha in Table 1. Alice has a negative efficacy-monitoring association.
The contrast shows why one pooled coefficient cannot represent every learner.
The confidence intervals quantify sampling uncertainty for each person and
pair.

## Plot the person-specific coefficients

The coefficient-and-interval display preserves the sign, magnitude, and
uncertainty of every association. Direct pair labels prevent the three
correlations for a learner from being mistaken for repeated estimates of one
quantity.

```{r correlation-plot, fig.cap="Figure 1. Distribution of within-person correlations for two learners."}
correlation_labels <- paste(correlations$subject,
                            paste(correlations$x, correlations$y, sep = " × "),
                            sep = " · ")
forest_plot(correlations$r, correlations$conf_low, correlations$conf_high,
            correlation_labels, xlab = "Within-person correlation (95% CI)")
```

Figure 1 shows that all three displayed associations are positive for Aisha.
Alice's efficacy-effort association is positive, while both correlations
involving monitoring are negative. The intervals make the disagreement
between people visible; the figure does not identify a causal direction.

## Assumptions and failure checks

`correlate_persons()` assumes paired numeric observations and an approximately
linear association. The estimate can be unstable when the number of complete
pairs is small. The `min_n` argument sets the minimum pair count. Constant
variables return missing correlations because their standard deviations are
zero.

`correlate_persons()` reports contemporaneous association. Serial dependence
can make ordinary correlation intervals optimistic because adjacent
observations may not be independent. Lagged prediction requires a temporal
model such as `fit_lm()` with lagged predictors or one of the network
estimators. Correlation alone does not adjust for other variables.

## When to use which

`correlate_persons()` is appropriate for an initial pairwise account of
within-person association. `fit_lm()` is appropriate when one outcome is
conditional on several predictors. `fit_within_between()` is appropriate when
within-person and between-person effects must be estimated separately.
Dynamic-network models are appropriate when several variables act jointly as
outcomes over time.

## References
