---
title: "Person-Level Description with describe_persons()"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Person-Level Description with describe_persons()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE, fig.width = 7.2, fig.height = 4.6, dpi = 144, fig.align = "center", out.width = "100%")
source("figure-style.R")
library(idiographic)
data(srl)
analysis_data <- subset(srl, name %in% c("Aisha", "Alice", "Anika", "Astrid", "Bjorn", "Bob", "Charlie", "Diana", "Erik", "Eve", "Fatima", "Frank"))
```

## Analytic question

`describe_persons()` answers whether each person's repeated series contains
enough variation, temporal coverage, and usable observations for an
idiographic analysis. The function takes a long-format panel, variable names,
a person identifier, and an optional ordering variable. It returns one tidy
row per person and variable.

`describe_persons()` separates overall dispersion from successive change.
The standard deviation measures dispersion around a person's mean. The root
mean square successive difference measures movement between adjacent
occasions. These quantities describe different temporal patterns
[@jahng2008]. The lag-one autocorrelation measures carry-over from one occasion
to the next.

## Estimate person-level summaries

`describe_persons()` uses `time = "day"` to order observations within each
learner. The `detail = "full"` setting adds floor and ceiling proportions,
acute-change probability, skewness, kurtosis, linear trend, and the longest
run of repeated values. Table 1 reports two learners and two variables so the
full output remains readable.

```{r description-table}
descriptives <- describe_persons(analysis_data, id = "name", vars = c("efficacy", "effort"), time = "day", subject = c("Aisha", "Alice"), detail = "full")
descriptives
```

`describe_persons()` reports 156 observations for each displayed series in
Table 1. Aisha's efficacy mean is higher than Alice's, while their efficacy
RMSSD values are similar. The result therefore distinguishes a difference in
typical level from a difference in occasion-to-occasion movement. The gap
columns equal one day for these learners, so a lag represents a consistent
elapsed interval.

## Plot dispersion against successive change

The figure compares standard deviation with RMSSD and labels every series
directly. Position, colour, and symbol jointly identify the variable and
learner; the labels make the figure interpretable without a separate lookup.

```{r description-plot, fig.cap="Figure 1. Overall dispersion and successive change for efficacy and effort in two learners."}
point_col <- ifelse(descriptives$variable == "efficacy", fig_col["blue"], fig_col["orange"])
point_pch <- ifelse(descriptives$subject == "Aisha", 21, 24)
labels <- paste(descriptives$subject, descriptives$variable, sep = " · ")
figure_begin(c(4.2, 4.8, 1, 2.2))
plot(descriptives$sd, descriptives$rmssd,
     xlim = range(descriptives$sd) + c(-1, 1) * diff(range(descriptives$sd)) * 0.12,
     ylim = range(descriptives$rmssd) + c(-1, 1) * diff(range(descriptives$rmssd)) * 0.12,
     pch = point_pch, bg = point_col, col = "white", cex = 1.5,
     xlab = "Person-level standard deviation",
     ylab = "Root mean square successive difference")
figure_grid(x = TRUE, y = TRUE)
points(descriptives$sd, descriptives$rmssd, pch = point_pch,
       bg = point_col, col = "white", cex = 1.5)
text(descriptives$sd, descriptives$rmssd, labels = labels, pos = 4,
     offset = 0.65, cex = 0.76, col = fig_col["ink"], xpd = TRUE)
```

Figure 1 shows that overall spread and short-term movement do not occupy one
common scale. A series can vary widely across the observation period without
changing by the same amount between adjacent days. Both quantities should be
reported when temporal instability is substantively relevant.

## Assumptions and failure checks

`describe_persons()` treats successive rows as adjacent only after ordering
within person. The `time` argument is therefore required when row order is not
the intended chronology. Missing values reduce the usable pair count for
RMSSD and autocorrelation. A constant series has no estimable correlation and
cannot support a person-specific slope.

`describe_persons()` defines acute change relative to the pooled sample unless
`pac_cutoff` is supplied. Analyses that compare acute-change rates across
datasets should use a common substantive cutoff. The `trend` and `trend_p`
columns describe a linear trend. They do not diagnose every form of
nonstationarity.

## When to use which

`describe_persons()` is the first choice for person-level data sufficiency,
location, variability, movement, and timing. `correlate_persons()` addresses
pairwise within-person association. `variance_components()` quantifies how
much sample variance lies within and between people. Model-fitting functions
should follow these checks when the available variation supports the intended
estimand.

## References
