---
title: "Discovering Person Subgroups with find_subgroups()"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Discovering Person Subgroups with find_subgroups()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE, fig.width = 7.2, fig.height = 4.6, dpi = 144, fig.align = "center", out.width = "100%")
source("figure-style.R")
library(idiographic)
data(srl)
set.seed(20260902)
analysis_data <- subset(srl, name %in% c("Aisha", "Alice", "Anika", "Astrid", "Bjorn", "Bob", "Charlie", "Diana", "Erik", "Eve", "Fatima", "Frank"))
```

## Analytic question

`find_subgroups()` assigns people to subgroups based on person-specific
effects, prediction errors, repeated splits, moderators, or mixtures of
regressions. The function takes a repeated-measures regression problem, a
method, a group count, and a resampling count. It returns group assignments and
assignment stability for each person.

`find_subgroups()` uses effect clustering here. The method estimates each
learner's regression coefficients and clusters those coefficient vectors.
Repeated resamples assess whether each learner returns to the same group.
Stability measures reproducibility of the partition. It does not test whether
separate populations exist.

## Estimate a two-group partition

`find_subgroups()` clusters efficacy and monitoring effects into two groups.
Twenty resamples provide an assignment-stability estimate.

```{r subgroup-find}
subgroup_result <- find_subgroups(analysis_data, y = "effort", x = c("efficacy", "monitoring"), id = "name", method = "effect_clustering", k = 2, reps = 20, time = "day")
subgroup_result
```

`find_subgroups()` returns two groups and reports their sizes. The print method
also states that existence was not tested because `k = 2` was supplied. A
stable forced partition can occur in a population without discrete classes.

## Inspect assignments

`groups()` returns one row per learner with the subgroup, method, stability,
and number of assignments. Table 1 sorts through the public `sort_by`
argument.

```{r subgroup-groups}
groups(subgroup_result, sort_by = "subgroup")
```

`groups()` shows which learners share a coefficient pattern in Table 1.
Stability values near one indicate consistent reassignment across resamples.
Values near the two-group chance level of 0.5 indicate uncertain membership.

## Plot assignment stability

`plot_subgroups()` colours learners by subgroup and draws the chance-level
reference. Figure 1 orders learners within subgroup by stability.

```{r subgroup-find-plot, fig.cap="Figure 1. Resampling stability of each learner's two-group assignment."}
plot_subgroups(subgroup_result)
```

Figure 1 identifies learners whose assignments are stable and learners near
the chance reference. Uncertain memberships should not be used as fixed labels
without sensitivity analysis.

## Assumptions and failure checks

`find_subgroups()` assumes that its feature representation corresponds to the
scientific definition of similarity. Effect clustering groups coefficient
patterns. Error clustering groups prediction-error patterns. Model trees use
person-level moderators. These methods answer different questions and can
produce different partitions.

`find_subgroups()` requires enough people per requested group and enough
observations to estimate person features. Coefficient skewness can create
apparent classes [@bauer2003]. `k = "auto"` delegates the group count to
`test_subgroups()` and can return one population.

## When to use which

`find_subgroups()` is appropriate for estimating assignments after the target
representation and group count are justified. `test_subgroups()` is
appropriate for deciding whether more than one population is supported.
`fit_subgroups()` is appropriate for evaluating whether group-specific models
improve held-out prediction.

## References
