---
title: "Pooled and Person-Specific Linear Models with fit_lm()"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Pooled and Person-Specific Linear Models with fit_lm()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE, fig.width = 7.2, fig.height = 4.6, dpi = 144, fig.align = "center", out.width = "100%")
source("figure-style.R")
library(idiographic)
data(srl)
analysis_data <- subset(srl, name %in% c("Aisha", "Alice", "Anika", "Astrid", "Bjorn", "Bob", "Charlie", "Diana", "Erik", "Eve", "Fatima", "Frank"))
```

## Analytic question

`fit_lm()` estimates a continuous outcome from one or more predictors at
pooled, subgroup, or person-specific scopes. The function takes a long-format
panel and column names for the outcome, predictors, person, and optional time
order. It returns an `idiographic_fit` with tidy coefficients, held-out
predictions, performance metrics, and failures.

`fit_lm()` uses an ordered hold-out within each learner. Earlier observations
form the training set and the final proportion forms the test set. The pooled
model shares one coefficient vector across learners. The individual scope fits
one coefficient vector per learner.

## Fit both scopes

`fit_lm()` models daily effort from efficacy and monitoring. The default
`scope = "both"` fits pooled and individual models using the same ordered test
rows. Table 1 compares their aggregate held-out performance.

```{r lm-fit}
lm_fit <- fit_lm(analysis_data, y = "effort", x = c("efficacy", "monitoring"), id = "name", time = "day", scope = "both")
metrics(lm_fit, overall = TRUE)
```

`fit_lm()` gives the individual models a lower RMSE than the pooled model in
Table 1. The individual aggregate RMSE is about 19.8, compared with 23.2 for
the pooled model. The corresponding held-out R-squared values are about 0.48
and 0.29. These figures describe this panel and this split.

## Inspect person-specific coefficients

`coefs()` retrieves inferential estimates without accessing the fit's internal
list. Table 2 reports Aisha's individual regression.

```{r lm-coefficients}
coefs(lm_fit, scope = "individual", subject = "Aisha")
```

`coefs()` reports positive efficacy and monitoring coefficients for Aisha in
Table 2. Each coefficient is conditional on the other predictor. The standard
errors and p-values describe uncertainty in the training-period regression.
They do not measure held-out prediction error.

## Plot held-out predictions

`plot_predictions()` compares observed and predicted test values for selected
learners. Figure 1 displays three individual models. Black points and lines are
observations. Orange lines are predictions.

```{r lm-plot, fig.cap="Figure 1. Held-out effort and person-specific linear-model predictions for three learners."}
plot_predictions(lm_fit, scope = "individual", n_subjects = 3)
```

Figure 1 shows that prediction quality differs across learners and occasions.
The aggregate advantage in Table 1 does not imply equal accuracy for every
person. The trajectory plot should accompany aggregate metrics when the unit
of inference is the person.

## Assumptions and failure checks

`fit_lm()` assumes a linear conditional mean, finite residual variance, and a
training period that represents the test period. Coefficient intervals also
use the fitted linear-model sampling assumptions. Serially correlated
residuals can make those intervals too narrow. Held-out metrics remain useful
for prediction even when coefficient inference requires a richer error model.

`fit_lm()` records persons with insufficient complete training or test rows in
the failure table. The function does not move observations across person
boundaries to fill a split. The optional `weights` argument names a case-weight
column. The `estimator = "robust"` option uses `MASS::rlm()` when resistance to
outlying residuals is required.

## When to use which

`fit_lm()` is appropriate for a continuous outcome and interpretable linear
coefficients. `fit_glm()` is appropriate for binary, count, and other supported
response families. `fit_ml()` is appropriate when held-out prediction and
nonlinear algorithms are primary. `fit_within_between()` is appropriate when
raw predictors would mix within-person and between-person effects.

## References
