---
title: "Within- and Between-Person Effects with fit_within_between()"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Within- and Between-Person Effects with fit_within_between()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE, fig.width = 7.2, fig.height = 4.6, dpi = 144, fig.align = "center", out.width = "100%")
source("figure-style.R")
library(idiographic)
data(srl)
analysis_data <- subset(srl, name %in% c("Aisha", "Alice", "Anika", "Astrid", "Bjorn", "Bob", "Charlie", "Diana", "Erik", "Eve", "Fatima", "Frank"))
```

## Analytic question

`fit_within_between()` estimates two coefficients for each time-varying
predictor. The within-person coefficient asks whether a person's outcome is
higher when the predictor is above that person's usual level. The
between-person coefficient asks whether people with higher predictor means
also have higher outcome means.

`fit_within_between()` takes a repeated-measures panel and returns an
`idiographic_wb` fit with component-labelled coefficients, predictions,
metrics, contextual contrasts, and variance components. The decomposition
follows the person-mean approach to panel models [@mundlak1978].

## Fit the hybrid model

`fit_within_between()` models effort from efficacy and monitoring at the
pooled scope. Table 1 reports within-person, between-person, and contextual
effects. The contextual effect is the between-person coefficient minus the
within-person coefficient.

```{r wb-fit}
wb_fit <- fit_within_between(analysis_data, y = "effort", x = c("efficacy", "monitoring"), id = "name", time = "day", scope = "pooled")
contextual(wb_fit)
```

`fit_within_between()` estimates a positive within-person efficacy coefficient
of about 0.36 and a positive between-person coefficient of about 0.57 in Table
1. Their contextual difference is about 0.21, and its confidence interval
includes zero. Monitoring has a positive within-person estimate and a negative
between-person estimate, but the contextual interval also includes zero.

## Inspect all coefficient components

`coefs()` retains the component and source-variable labels. Table 2 shows the
pooled coefficient table.

```{r wb-coefficients}
coefs(wb_fit, scope = "pooled")
```

`coefs()` identifies efficacy's within-person coefficient as statistically
different from zero in Table 2. The estimate describes daily deviations from a
learner's own efficacy mean. It should not be reported as a contrast between
learners.

## Plot within and between estimates

`plot_components()` draws component estimates and confidence intervals for
each predictor. Figure 1 uses orange for within-person effects, blue for
between-person effects, and green for contextual differences.

```{r wb-plot, fig.cap="Figure 1. Within-person, between-person, and contextual coefficients with confidence intervals."}
plot_components(wb_fit, scope = "pooled")
```

Figure 1 shows that efficacy has positive within-person and between-person
coefficients. Monitoring changes sign across levels. The intervals convey
greater uncertainty in the between-person and contextual quantities because
they depend on 12 independent learner means.

## Assumptions and failure checks

`fit_within_between()` requires predictors that vary within people and across
people. A person-constant predictor cannot identify a person-specific within
effect. A predictor with identical person means cannot identify a between
effect. The OLS estimator clusters inference by person. The optional mixed
estimator delegates to `lme4`.

`fit_within_between()` gives distinct names to decomposed columns and labels
each returned coefficient by component. This prevents a raw-score coefficient
from being interpreted at the wrong level. The model remains associational
unless the research design and covariate assumptions identify a causal effect.

## When to use which

`fit_within_between()` is appropriate when a predictor varies at both levels
and the distinction is substantively important. `preprocess_panel()` is
appropriate when decomposed columns are needed for another modelling function.
`fit_lm()` is appropriate when the predictor is already defined at the intended
level or when a raw conditional association is the target.

## References
