---
title: "Treatment-Effect Estimation with fit_effects()"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Treatment-Effect Estimation with fit_effects()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE, fig.width = 7.2, fig.height = 4.6, dpi = 144, fig.align = "center", out.width = "100%")
source("figure-style.R")
library(idiographic)
data(srl)
analysis_data <- subset(srl, name %in% c("Aisha", "Alice", "Anika", "Astrid", "Bjorn", "Bob", "Charlie", "Diana", "Erik", "Eve", "Fatima", "Frank"))
analysis_data <- transform(analysis_data, high_monitoring = as.integer(monitoring > median(monitoring, na.rm = TRUE)))
```

## Analytic question

`fit_effects()` estimates average and heterogeneous effects for binary,
multi-arm, or continuous treatments. The function takes an outcome, treatment,
covariates, person identifier, optional time order, and model controls. It
returns an `idiographic_effects` object with tidy effect estimates,
predictions, metrics, and failures.

`fit_effects()` uses augmented inverse-probability weighting for binary and
multi-arm treatments [@bang2005]. Outcome and propensity models are estimated
on training rows and applied to held-out rows. Continuous treatments use a
partially linear residualization score [@chernozhukov2018].

## Define the illustrative contrast

`fit_effects()` treats days above the median monitoring score as the exposed
condition. The bundled data are observational and contain no randomized
treatment. The example therefore illustrates estimation and diagnostics. It
does not identify a causal effect of monitoring.

`fit_effects()` adjusts for efficacy, planning, and control at the pooled
scope. Three sorted effect groups summarize estimated heterogeneity.

```{r effects-fit}
effect_fit <- fit_effects(analysis_data, y = "effort", treatment = "high_monitoring", x = c("efficacy", "planning", "control"), id = "name", time = "day", scope = "pooled", n_groups = 3)
effects(effect_fit)
```

`effects()` reports an average contrast of about -3.7 in Table 1. Its
confidence interval includes zero. The top-minus-bottom sorted-group contrast
is about 15.1, and its interval also includes zero. The BLP heterogeneity
interval includes zero. This analysis does not provide evidence of a stable
effect difference across the sorted groups.

## Plot sorted effect groups

`plot_effects()` draws each GATES estimate with its confidence interval.
Figure 1 orders groups from the lowest to highest predicted effect and marks
zero with an orange reference line.

```{r effects-plot, fig.cap="Figure 1. Sorted high-monitoring contrasts and confidence intervals."}
plot_effects(effect_fit, scope = "pooled")
```

Figure 1 shows overlapping intervals across the three groups. The ordering was
learned from the data, while the displayed group estimates use held-out scores.
Wide intervals reflect uncertainty across 12 learners.

## Identification assumptions and failure checks

`fit_effects()` requires consistency, positivity, and conditional exchangeability
for a causal interpretation. Positivity requires both treatment conditions
across relevant covariate patterns. Conditional exchangeability requires that
the supplied covariates block treatment-outcome confounding. The observational
teaching contrast cannot verify that condition.

`fit_effects()` removes rows with a missing treatment before it constructs the
temporal split. Missing outcomes and predictors are handled by the shared
complete-case split. A person can still enter the failure table when too few
usable training or test rows remain.

`fit_effects()` trims estimated propensities away from zero and one. Severe
trimming indicates poor overlap and changes the effective target population.
AIPW tolerates misspecification of one nuisance model under its regularity
conditions. Misspecifying both outcome and propensity models can bias the
estimate.

## When to use which

`fit_effects()` is appropriate when the target is an average or
person-varying treatment effect and the design supports causal identification.
`fit_lm()` is appropriate for conditional association. `fit_heterogeneity()`
is appropriate for repeated-split inference about treatment effects,
prediction error, or the gain from individual modelling.

## References
