---
title: "Generalized Linear Models with fit_glm()"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Generalized Linear Models with fit_glm()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE, fig.width = 7.2, fig.height = 4.6, dpi = 144, fig.align = "center", out.width = "100%")
source("figure-style.R")
library(idiographic)
data(srl)
analysis_data <- subset(srl, name %in% c("Aisha", "Alice", "Anika", "Astrid", "Bjorn", "Bob", "Charlie", "Diana", "Erik", "Eve", "Fatima", "Frank"))
analysis_data <- transform(analysis_data, high_effort = factor(ifelse(effort > median(effort, na.rm = TRUE), "high", "low")))
```

## Analytic question

`fit_glm()` estimates generalized linear models at pooled, subgroup, and
person-specific scopes. The function accepts the same panel structure as
`fit_lm()` plus a response family. It returns tidy held-out classification or
regression metrics, predictions, coefficients, and failures.

`fit_glm()` supports Gaussian, binomial, and Poisson families with base R. The
negative-binomial option delegates to `MASS::glm.nb()`. A binomial model uses a
logit link and returns probabilities for the positive outcome level
[@mccullagh1989].

## Define and model a binary outcome

`fit_glm()` models whether daily effort exceeds the median in the complete
12-learner panel. The derived outcome supports a reproducible classification
example. It is not a clinical or policy threshold. Table 1 compares pooled and
individual held-out performance.

```{r glm-fit}
glm_fit <- fit_glm(analysis_data, y = "high_effort", x = c("efficacy", "monitoring"), id = "name", family = "binomial", time = "day", scope = "both")
metrics(glm_fit, overall = TRUE)
```

`fit_glm()` gives the individual models an aggregate accuracy of about 0.74 in
Table 1. The pooled model has an accuracy of about 0.69. The individual models
also have lower Brier score and log loss. All three metrics favour the
individual scope for this outcome and split.

## Inspect the probability model

`coefs()` returns log-odds coefficients. Table 2 keeps the pooled scope so the
model has one coefficient vector across learners.

```{r glm-coefficients}
coefs(glm_fit, scope = "pooled")
```

`coefs()` reports changes in log odds per one-unit predictor increase in Table
2. Exponentiating a coefficient would produce an odds ratio, but the displayed
scale preserves the fitted model's direct parameterization. Held-out
probabilities should be assessed alongside coefficient inference.

## Plot classification calibration

`plot_diagnostics()` plots predicted probability against whether each held-out
classification was correct. Figure 1 adds a lowess curve to summarize the
relation.

```{r glm-plot, fig.cap="Figure 1. Held-out classification correctness across predicted probabilities."}
plot_diagnostics(glm_fit, scope = "pooled", type = "calibration")
```

Figure 1 assesses whether high reported probabilities correspond to more
correct classifications. The plot does not replace a calibration intercept or
slope. It identifies ranges where probability estimates warrant closer
inspection.

## Assumptions and failure checks

`fit_glm()` assumes that the chosen family and link describe the conditional
outcome distribution. A binomial outcome must contain exactly two observed
levels. The package selects the final sorted level as the positive class unless
factor levels specify an order. Poisson regression assumes equality of the
conditional mean and variance. The negative-binomial family is preferable when
count variance exceeds the mean.

`fit_glm()` can fail for a person when the training outcome contains one class,
when complete rows are too few, or when coefficients separate perfectly. The
failure table retains these cases. Aggregate metrics should be interpreted
with the number of successful people.

## When to use which

`fit_glm()` is appropriate when the outcome distribution has a supported
family and coefficient interpretation matters. `fit_lm()` is appropriate for
a continuous outcome with an identity link. `fit_ml()` is appropriate when
algorithm comparison and held-out prediction have priority over a single
parametric coefficient model.

## References
