---
title: "Repeated-Split Heterogeneity Analysis with fit_heterogeneity()"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Repeated-Split Heterogeneity Analysis with fit_heterogeneity()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE, fig.width = 7.2, fig.height = 4.6, dpi = 144, fig.align = "center", out.width = "100%")
source("figure-style.R")
library(idiographic)
data(srl)
set.seed(20260902)
analysis_data <- subset(srl, name %in% c("Aisha", "Alice", "Anika", "Astrid", "Bjorn", "Bob", "Charlie", "Diana", "Erik", "Eve", "Fatima", "Frank"))
```

## Analytic question

`fit_heterogeneity()` tests whether a person-varying quantity differs
systematically across people. The target can be a treatment effect, prediction
error, or gain from individual rather than pooled modelling. The function
returns repeated-split estimates, sorted groups, a heterogeneity slope,
characteristics of extreme groups, and learner-selection criteria.

`fit_heterogeneity()` separates proxy learning from inference. One sample
learns a machine-learning proxy for the person-varying target. A held-out
sample estimates average, grouped, and heterogeneity quantities. Repeating and
aggregating the split reduces dependence on one random partition
[@chernozhukov2020]. Standard errors cluster observations by person.

## Study heterogeneity in prediction error

`fit_heterogeneity()` asks whether effort prediction error varies predictably
across learners. Linear and ridge proxies compete over ten person-level random
splits. Three sorted groups range from lower to higher error.

```{r heterogeneity-fit}
heterogeneity_fit <- fit_heterogeneity(analysis_data, y = "effort", x = c("efficacy", "monitoring", "planning"), id = "name", target = "error", model = c("linear", "ridge"), num_splits = 10, n_groups = 3)
heterogeneity(heterogeneity_fit)
```

`heterogeneity()` reports an average absolute prediction error of about 20.2
in Table 1. The highest-minus-lowest group contrast is about 3.0, and its
interval includes zero. The heterogeneity slope interval also includes zero.
The analysis does not find stable, predictable error heterogeneity in this
12-learner subset.

## Describe and compare learned proxies

`clan()` compares predictor means between extreme target groups. `learners()`
reports the criterion used to select the proxy model.

```{r heterogeneity-details}
clan(heterogeneity_fit)
learners(heterogeneity_fit)
```

`clan()` reports descriptive differences between extreme predicted-error
groups in Table 2. These differences describe the fitted grouping and do not
identify causes of prediction error. `learners()` selects ridge by a small
criterion difference in Table 3.

## Plot the heterogeneity estimates

The coefficient-and-interval display includes the average, heterogeneity
slope, sorted groups, and extreme-group contrast. Unlike a point-only plot, it
keeps the uncertainty required for the substantive conclusion.

```{r heterogeneity-plot, fig.cap="Figure 1. Repeated-split estimates for average error and error heterogeneity."}
heterogeneity_table <- heterogeneity(heterogeneity_fit)
heterogeneity_labels <- c(
  average = "Average absolute error",
  heterogeneity = "Heterogeneity slope",
  `group:g1` = "Lowest-error group",
  `group:g2` = "Middle-error group",
  `group:g3` = "Highest-error group",
  `group:top-bottom` = "Highest minus lowest"
)
labels <- unname(heterogeneity_labels[heterogeneity_table$effect])
forest_plot(heterogeneity_table$estimate, heterogeneity_table$conf_low,
            heterogeneity_table$conf_high, labels,
            xlab = "Absolute-error target (95% CI)", colour = fig_col["purple"])
```

Figure 1 shows similar estimates for the three error groups and wide
uncertainty for their contrast. The heterogeneity-slope and highest-minus-lowest
intervals cross zero, so this analysis does not establish systematic error
heterogeneity.

## Assumptions and failure checks

`fit_heterogeneity()` treats people as independent clusters for the default
person split. The number of independent people limits precision even when each
person contributes many rows. More random splits reduce split sensitivity but
do not create additional independent units.

`fit_heterogeneity()` uses random rather than time-ordered splits because its
target is cross-person heterogeneity. The `gain` target changes to occasion
splits so each person appears in both modelling samples. Forecast evaluation
should use `fit_rolling()` instead. A causal interpretation of the `cate`
target requires the same identification assumptions as `fit_effects()`.

## When to use which

`fit_heterogeneity()` is appropriate for repeated-split inference about a
person-varying target. `fit_effects()` is appropriate for direct average and
sorted treatment-effect estimation across model scopes. `test_subgroups()` is
appropriate when the question concerns discrete coefficient populations.
`fit_rolling()` is appropriate for change in prediction performance over time.

## References
