---
title: "Subgroup-Specific Models with fit_subgroups()"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Subgroup-Specific Models with fit_subgroups()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE, fig.width = 7.2, fig.height = 4.6, dpi = 144, fig.align = "center", out.width = "100%")
source("figure-style.R")
library(idiographic)
data(srl)
set.seed(20260902)
analysis_data <- subset(srl, name %in% c("Aisha", "Alice", "Anika", "Astrid", "Bjorn", "Bob", "Charlie", "Diana", "Erik", "Eve", "Fatima", "Frank"))
```

## Analytic question

`fit_subgroups()` fits one model per discovered or supplied subgroup and can
compare those models with pooled and individual alternatives. The function
takes a panel, outcome, predictors, person identifier, subgroup mapping, and
model family. It returns an `idiographic_fit` with the usual tidy accessors.

`fit_subgroups()` defaults to `scope = "all"`. The pooled scope estimates one
coefficient vector. The subgroup scope estimates one vector per group. The
individual scope estimates one vector per person. All scopes predict the same
ordered test rows.

## Estimate subgroup assignments

`find_subgroups()` supplies a two-group effect-clustering result. This first
stage defines the subgroup mapping used by the fitted models.

```{r subgroup-stage}
subgroup_result <- find_subgroups(analysis_data, y = "effort", x = c("efficacy", "monitoring"), id = "name", method = "effect_clustering", k = 2, reps = 20, time = "day")
groups(subgroup_result, sort_by = "subgroup")
```

`groups()` reports assignments and their resampling stability in Table 1.
These labels are estimated quantities. Prediction comparisons should retain
that first-stage uncertainty in their interpretation.

## Fit and compare three scopes

`fit_subgroups()` fits linear models for all three scopes. Table 2 reports
aggregate held-out metrics.

```{r subgroup-fit}
subgroup_fit <- fit_subgroups(analysis_data, y = "effort", x = c("efficacy", "monitoring"), id = "name", subgroup = subgroup_result, method = "lm", time = "day")
metrics(subgroup_fit, overall = TRUE)
```

`fit_subgroups()` gives the subgroup models an RMSE of about 22.8 in Table 2.
The pooled RMSE is about 23.2, and the individual RMSE is about 19.8. The
two-group model improves modestly over pooling but does not match the
individual models on this split.

## Plot the scope comparison

`plot_metrics()` displays the common held-out metric for pooled, subgroup, and
individual scopes. Position and colour distinguish the modelling scopes.

```{r subgroup-fit-plot, fig.cap="Figure 1. Held-out RMSE for pooled, subgroup, and individual linear models."}
plot_metrics(subgroup_fit, metric = "rmse")
```

Figure 1 shows the subgroup model between the pooled and individual models in
prediction error. The result quantifies the compromise between one model for
everyone and one model per person.

## Assumptions and failure checks

`fit_subgroups()` treats the supplied mapping as fixed during model fitting.
Assignments estimated from the same outcome data can make subgroup performance
optimistic. A confirmatory analysis should derive groups in separate data or
repeat the complete discovery-and-fitting pipeline within resampling.

`fit_subgroups()` requires at least one usable training and test set for each
requested scope. Small groups can fail even when pooled and individual models
succeed. The failure table identifies these cases. Subgroup prediction does
not by itself show that discrete populations exist.

## When to use which

`fit_subgroups()` is appropriate when a defensible mapping exists and the
question concerns group-specific prediction. `fit_lm()` or `fit_glm()` with a
subgroup column provides the same lower-level fitting engine. `test_subgroups()`
addresses existence. `find_subgroups()` addresses assignment.

## References
