---
title: "Rolling-Origin Validation with fit_rolling()"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Rolling-Origin Validation with fit_rolling()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE, fig.width = 7.2, fig.height = 4.6, dpi = 144, fig.align = "center", out.width = "100%")
source("figure-style.R")
library(idiographic)
data(srl)
analysis_data <- subset(srl, name %in% c("Aisha", "Alice", "Anika", "Astrid", "Bjorn", "Bob", "Charlie", "Diana", "Erik", "Eve", "Fatima", "Frank"))
```

## Analytic question

`fit_rolling()` estimates forecasting performance across several points in
time. The function takes a panel, outcome, predictors, person and time columns,
a model family, and a rolling schedule. It returns the standard
`idiographic_fit` tables with a fold identifier added to predictions.

`fit_rolling()` trains on observations before each origin and predicts the next
assessment block. The origin then moves forward by `step` observations. This
expanding-window design evaluates a model under repeated historical forecasts
instead of one terminal hold-out [@tashman2000].

## Define the rolling schedule

`fit_rolling()` fits pooled linear models at four origins. Each fold predicts
the next five observations per learner. The default initial window uses all
earlier rows that can accommodate the requested schedule. Table 1 pools the
forecast errors across folds.

```{r rolling-fit}
rolling_fit <- fit_rolling(analysis_data, y = "effort", x = c("efficacy", "monitoring"), id = "name", method = "lm", time = "day", scope = "pooled", folds = 4, assess = 5)
metrics(rolling_fit, overall = TRUE)
```

`fit_rolling()` evaluates 240 forecasts in Table 1. This count equals four
folds, five test occasions, and 12 learners. The pooled linear model has an
RMSE of about 23.8 and a positive mean error of about 1.4. These metrics
summarize repeated forward predictions rather than in-sample fit.

## Inspect fold-level predictions

`predictions()` returns one row per forecast with the original row identifier,
person, observed value, predicted value, and fold. Table 2 limits printing
through the accessor's `n` argument.

```{r rolling-predictions}
predictions(rolling_fit, n = 12)
```

`predictions()` shows that each row belongs to a specific forecast origin in
Table 2. The fold column permits error review by historical period without
reaching into the fitted object.

## Plot forecast trajectories

`plot_predictions()` draws observed and predicted values in row order for
selected learners. Figure 1 uses three learner panels.

```{r rolling-plot, fig.cap="Figure 1. Observed effort and pooled rolling-origin predictions for three learners."}
plot_predictions(rolling_fit, scope = "pooled", n_subjects = 3)
```

Figure 1 shows where the pooled forecast follows or misses each learner's
held-out trajectory. The plot retains person labels even though one coefficient
vector is fitted at the pooled scope.

## Assumptions and failure checks

`fit_rolling()` requires ordered observations and a schedule that fits the
shortest person series. The function stops when `initial`, `assess`, `step`,
and `folds` require more rows than are available. Missing predictor or outcome
values can reduce usable rows within a fold.

`fit_rolling()` can tune machine-learning models inside each fold. Validation
rows are taken from that fold's training period. The assessment block remains
unseen during tuning. Overlapping assessment blocks occur when `step` is
smaller than `assess`; metrics then count some time regions more than once.

## When to use which

`fit_rolling()` is appropriate for repeated prospective evaluation. `fit_lm()`,
`fit_glm()`, and `fit_ml()` use one terminal hold-out and are appropriate for a
single train-test comparison. Rolling network functions address time-varying
multivariate systems and have a different result contract.

## References
