---
title: "Idiographic Machine Learning with fit_ml()"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Idiographic Machine Learning with fit_ml()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE, fig.width = 7.2, fig.height = 4.6, dpi = 144, fig.align = "center", out.width = "100%")
source("figure-style.R")
library(idiographic)
data(srl)
analysis_data <- subset(srl, name %in% c("Aisha", "Alice", "Anika", "Astrid", "Bjorn", "Bob", "Charlie", "Diana", "Erik", "Eve", "Fatima", "Frank"))
```

## Analytic question

`fit_ml()` asks a prospective question: how accurately can measurements
already observed predict a person's later outcome? It compares algorithms at
pooled, subgroup, or individual scopes and returns held-out predictions,
person-level and aggregate metrics, tuning results, coefficients or importance
where available, and explicit fitting failures.

This question differs from explaining an in-sample association. A flexible
model can reproduce its training data and still fail on later observations.
Accordingly, every comparison below uses the same chronologically later test
rows. The analysis is predictive; it does not estimate a causal effect of
efficacy, monitoring, or planning on effort.

## How temporal validation works

When `time` is supplied, observations are ordered within person before they
are assigned to three roles. The default proportions in this example imply the
following sequence for each learner.

| Data block | Used for | Visible to final metric? |
|:--|:--|:--|
| Earlier training rows | Estimate model parameters | No |
| Later validation rows | Select a hyperparameter when `tune = TRUE` | No |
| Final test rows | Estimate future predictive performance once | Yes |

After tuning, `fit_ml()` refits the selected setting on the combined training
and validation rows and predicts the untouched test block. Preprocessing is
also learned from training data rather than the test period. This nested
separation prevents test outcomes from choosing the model [@hastie2009]. It
does not remove temporal drift: if the process changes after the observed
period, even an honestly estimated test error can be optimistic.

## Inspect the available algorithms

`models()` is the public registry for algorithms and backends. Native models
require no optional package; the `parameter` column identifies what tuning
changes. A mean model is included because a complex method should be required
to outperform a defensible baseline.

```{r ml-models}
native_regression <- as.data.frame(models(task = "regression", available = TRUE))
subset(native_regression, estimator == "native",
       select = c("model", "estimator", "parameter"))
```

The native set spans a mean baseline, regularised and unregularised linear
models, principal-components regression, nearest neighbours, and a tree.
Optional backends add algorithms such as forests, support-vector machines, and
boosting when their packages are available. Algorithm breadth is useful only
when all candidates receive the same split and evaluation metric.

## Compare predictive algorithms

The first analysis compares five native regression algorithms for daily
effort. `scope = "pooled"` holds the modelling scope constant so that Figure 1
isolates algorithm choice. The predictors are efficacy, monitoring, and
planning measured on the same day; the target is therefore later held-out
rows, not a lagged causal response.

```{r ml-fit}
set.seed(20260902)
ml_fit <- fit_ml(
  analysis_data,
  y = "effort",
  x = c("efficacy", "monitoring", "planning"),
  id = "name",
  time = "day",
  scope = "pooled",
  model = c("mean", "linear", "ridge", "knn", "tree"),
  tune = TRUE
)
metrics(ml_fit, overall = TRUE)
```

Table 1 shows that k-nearest neighbours has the lowest test RMSE, about 9.2.
The mean baseline has an RMSE of about 27.6; linear and ridge regression both
reach about 21.9; the tree reaches about 25.4. The nearest-neighbour result is
therefore a large improvement over both the baseline and the linear
specifications for this panel. It is not a universal ranking of algorithms.

```{r ml-performance-plot, fig.cap="Figure 1. Held-out error for five pooled prediction algorithms. Lower RMSE is better."}
plot_metrics(ml_fit, metric = "rmse")
```

Figure 1 uses a common zero origin and direct values, which makes the size of
the differences auditable. The comparison is fair because the models predict
the same 384 test rows. Repeated response patterns in this teaching data can
favour a local-neighbour method; an independent dataset may not preserve that
advantage.

## Audit the tuning decision

`best_model()` ranks models with the task-appropriate metric. `tuning()`
reports the validation result for every candidate without exposing backend
objects. The tuning table is evidence about model selection; it is not the
final test evaluation.

```{r ml-selection}
best_model(ml_fit)
tuning(ml_fit, model = "knn")
```

`best_model()` selects k-nearest neighbours in Table 2. The tuning results show
that one neighbour has the smallest validation RMSE among the searched values.
Larger neighbourhoods smooth more aggressively and perform worse in this
example.

```{r ml-tuning-plot, fig.cap="Figure 2. Validation error across candidate neighbourhood sizes. The orange point is the selected value."}
plot_tuning(ml_fit, model = "knn", metric = "rmse")
```

Figure 2 exposes the full selection path. Reporting only the winning value
would hide whether the decision was decisive or nearly tied. The grid should
be defined from plausible values before inspecting the test results; it can be
overridden explicitly through `grid`.

```{r ml-custom-grid, eval=FALSE}
fit_ml(
  analysis_data,
  y = "effort", x = c("efficacy", "monitoring", "planning"),
  id = "name", time = "day", scope = "pooled", model = "knn",
  tune = TRUE, grid = list(knn = list(k = c(1, 3, 5, 9, 15, 25)))
)
```

## Decide whether pooling helps

Algorithm and scope answer separate questions. The next fit holds the winning
algorithm constant and compares one pooled model with one model per learner.
Both scopes retain the same final proportion of each person's timeline.

```{r ml-scope}
set.seed(20260902)
scope_fit <- fit_ml(
  analysis_data,
  y = "effort", x = c("efficacy", "monitoring", "planning"),
  id = "name", time = "day", scope = "both", model = "knn",
  tune = TRUE
)
metrics(scope_fit, overall = TRUE)
```

The pooled model has a lower aggregate RMSE than the individual models in
Table 4. Individualisation is not automatically superior: estimating twelve
separate neighbourhood structures reduces the training information available
to each model. The result supports partial or full pooling for this dataset,
although individual performance still requires inspection.

```{r ml-scope-plot, fig.cap="Figure 3. Held-out RMSE for pooled and individual k-nearest-neighbour models."}
plot_metrics(scope_fit, metric = "rmse")
```

Figure 3 isolates scope choice from algorithm choice. A subgroup analysis
would add `scope = "subgroup"` and a subgroup vector; subgroup discovery and
subgroup fitting should use separate data or resampling to avoid evaluating a
partition on the observations that created it.

## Find the people the winning model misses

An acceptable aggregate score can conceal poor predictions for particular
people. `metrics()` returns one row per person as well as `.overall`. The next
table orders the pooled k-nearest-neighbour errors from largest to smallest.

```{r ml-person-metrics}
person_error <- metrics(
  ml_fit, model = "knn", scope = "pooled",
  sort_by = "rmse", decreasing = TRUE
)
person_error <- subset(person_error, subject != ".overall")
head(person_error, 5)
```

```{r ml-person-plot, fig.cap="Figure 4. Person-level held-out RMSE for the selected pooled model. The dashed line is the aggregate RMSE."}
plot_subjects(ml_fit, model = "knn", scope = "pooled")
```

Figure 4 shows substantial variation around the aggregate result. Anika,
Frank, and Aisha have the largest errors, whereas the repeated patterns for Eve
and Erik are predicted almost exactly. This distribution is a core
idiographic result: the model's average utility does not guarantee utility for
every person.

## Inspect failures in time order

`plot_predictions()` compares observed and predicted test trajectories for the
three people with the largest RMSE. Selection is based only on the public
person-level metrics table.

```{r ml-trajectories, fig.height=7.2, fig.cap="Figure 5. Observed and predicted held-out effort for the three learners with the largest RMSE."}
hardest_people <- head(person_error$subject, 3)
plot_predictions(
  ml_fit, model = "knn", scope = "pooled",
  subject = hardest_people, n_subjects = 3
)
```

Figure 5 distinguishes a persistent level error from an isolated miss and
from failure to track changes. RMSE alone cannot make that distinction. A
trajectory plot should therefore accompany the aggregate metric whenever the
timing of errors matters.

## Explain predictions without treating importance as causality

K-nearest neighbours has no regression coefficient. Permutation importance
measures how much held-out RMSE increases when one predictor is shuffled while
the fitted model is kept fixed. `plot_importance()` averages this increase
across people for the pooled fit.

```{r ml-importance, fig.cap="Figure 6. Model-agnostic permutation importance for the selected pooled model."}
set.seed(20260902)
plot_importance(
  ml_fit, model = "knn", scope = "pooled", n = 3,
  method = "permutation", repeats = 20
)
```

Figure 6 ranks predictors by their contribution to held-out prediction error.
Importance has no sign and is not an intervention effect. Correlated
predictors can substitute for one another, making each appear less important
than the information they provide jointly.

## Assumptions and failure checks

The analysis depends on five checks.

1. **Temporal target.** Same-day predictors answer a different question from
   lagged predictors. Create lags with `preprocess_panel()` when the intended
   target is a future change conditional on earlier measurements.
2. **Comparable rows.** Changing the predictor set can change which complete
   rows enter a fit. Compare candidate models using one prespecified set of
   required variables.
3. **Sufficient information.** Individual models need enough complete training,
   validation, and test rows per person. Failed units are retained in the fit's
   failure table rather than silently removed.
4. **Non-stationarity.** A final holdout represents one historical transition.
   Use `fit_rolling()` when performance must be assessed across several forecast
   origins or when drift is itself substantively important.
5. **Selection uncertainty.** Small validation differences can reverse under a
   new split. Inspect `tuning()` and, for important decisions, repeat the
   analysis across defensible origins rather than reporting only the winner.

For regression, RMSE emphasises large errors, MAE gives equal linear weight to
errors, bias detects systematic over- or under-prediction, and test-set
R-squared compares the model with the test-set mean. Classification adds
accuracy, Brier score, and log loss. The metric should be chosen from the cost
of errors, not from whichever column makes a model appear best.

## When to use which

Use `fit_ml()` when the primary estimand is future predictive performance and
algorithm flexibility is substantively acceptable. Use `fit_lm()` or
`fit_glm()` when a prespecified coefficient and its uncertainty are the main
result. Use `fit_rolling()` when one final test block is insufficient. Use
`fit_subgroups()` when a defensible external or separately discovered grouping
defines the modelling scope. Use `fit_effects()` when the question is what an
intervention changes rather than what predicts an outcome.

A reproducible report should name the outcome and prediction horizon, list the
predictors available at prediction time, describe the person-wise split,
identify the selection metric and tuning grid, report the baseline and all
candidate test metrics, show the distribution of person-level performance,
and state any failed units. Figures 1--6 provide that audit trail for the
worked example.

## References
