---
title: "8. GIMME"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{8. GIMME}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 5.5
)
library(idiographic)
data(srl)
has_cograph <- requireNamespace("cograph", quietly = TRUE)
```

Group iterative multiple model estimation (GIMME) sits at the boundary of
idiographic and group inference. It estimates a person-specific unified
structural equation model for each individual — one structural model per
person, fitted to that person's own ordered series — while searching for
paths that recur in a sufficient proportion of individuals and promoting
those to a shared group level. The estimand is
therefore double: every subject receives an idiographic dynamic model, and
the group model records which pieces of structure the sample holds in
common. Like the other lag-one models in this package, it presumes weakly
stationary series, linear lag-one dynamics, and correctly ordered,
approximately equally spaced occasions within each person.

Two kinds of within-person path enter the search. A temporal path
`from -> to` is a directed lag-one path within a person's series: the value
of `from` at occasion $t-1$ predicts the value of `to` at occasion $t$,
holding the other lagged variables constant. A contemporaneous path is a
directed within-occasion path: `from` predicts `to` at the same occasion,
over and above what the lagged variables explain. Unlike the graphical VAR,
whose contemporaneous layer is an undirected partial-correlation network,
GIMME places directed SEM paths within an occasion as well as across
occasions. The weight displayed on a GIMME edge is the proportion of
subjects whose individual model carries that path — path prevalence — not a
regression coefficient; a weight of 1 states that every subject has the
path and by itself says nothing about the size or sign of the effect. The
per-person coefficients live in the individual models and are reported
separately.

This placement distinguishes GIMME from its neighbours. Multilevel VAR
(`fit_mlvar()`) pools all subjects into a single fixed-effect average
temporal matrix and treats person-level departures as random effects around
it; GIMME instead keeps one structural model per person and asks which
paths replicate. Unified SEM (`fit_usem()`) fits the same person-specific
model class for a single subject, with no group search; GIMME is the
multi-subject extension that adds the replication rule. The method is
appropriate when theory expects both shared pathways and person-specific
deviations, and when the question is which is which.

# Data and preprocessing

The estimator expects long format: one row per person-occasion, an id
column, an ordering column, and numeric time-varying indicators. The
bundled `srl` data hold self-regulated-learning indicators for 36 students
measured over 156 occasions each; `name` identifies the student and `day`
orders the occasions. Five indicators enter the model: `efficacy`, `value`,
`planning`, `monitoring`, and `effort`. Because GIMME estimates a full SEM
per subject, each individual series must carry the model on its own, so the
screen below also requires at least 100 usable occasions per student.

```{r vars-audit}
vars <- c("efficacy", "value", "planning", "monitoring", "effort")
preprocess(srl, vars = vars, id = "name", min_obs = 100)
```

The 36 students supply 5616 ordered rows, of which 5548 survive as complete
current/lagged pairs. Ten of the 180 subject-series trip the linear-trend
flag and one shows drift; none shows high autoregression, unit-root risk,
or zero variance. The fitted example uses eight students to keep the
vignette fast while preserving the `srl` measurement design.

# Fitting the model

The estimator takes the data, the variable set, the id column, and the time
column; `time = "day"` orders occasions within each student. `ar = TRUE`
places an autoregressive path on every variable in every subject's model
from the outset, which anchors the search. `groupcutoff = 0.75` and
`subcutoff = 0.75` require a path to improve fit in at least 75 percent of
the relevant subjects — here, six of the eight — before it is promoted by
the corresponding rule.

```{r fit}
vars <- c("efficacy", "value", "planning", "monitoring", "effort")
students <- subset(srl, name %in% c("Grace", "Eve", "Aisha", "Alice",
                                    "Bob", "Diana", "Frank", "Heidi"))
gimme_fit <- fit_gimme(
  students, vars = vars, id = "name", time = "day",
  ar = TRUE, groupcutoff = 0.75, subcutoff = 0.75, seed = 1
)
gimme_fit
```

The group model contains the five autoregressive paths, fixed by
`ar = TRUE` and carried by all eight students, plus the two contemporaneous
paths the group search added: efficacy predicting monitoring and planning
predicting effort, each present in every student's model. No cross-variable
temporal path met the group criterion, so the shared lag-one structure is
autoregressive only — a statement about replication at these cutoffs, not
about the absence of lagged effects in any one student. The
individual-level search then added between one and four further
contemporaneous paths per student, 2.5 on average. The printed matrices
hold prevalences: the temporal matrix has 1 on its diagonal and 0
elsewhere, and the contemporaneous matrix runs from 0.125 — a path carried
by a single student — up to 1 for the two group paths.

# Reading the output

The `summary()` method reports one row per network layer, counting
cross-variable edges only.

```{r summary}
summary(gimme_fit)
```

The temporal layer shows zero edges and density 0, because its only
retained paths are the self-loops, which are tallied separately. The
contemporaneous layer holds 13 directed edges at density 0.65 — 13 of the
20 possible ordered pairs — with a mean prevalence of 0.346 across retained
paths. The `edges()` accessor lists every retained path with its layer,
prevalence, and level.

```{r edges}
edges(gimme_fit)
```

The five temporal rows are the autoregressive self paths at prevalence 1,
labelled group. The contemporaneous rows split by level: the two group
paths at prevalence 1, then eleven individual-level paths, from efficacy
predicting planning and monitoring predicting value at 0.375 — three of the
eight students — down to four paths carried by a single student at 0.125.

```{r coefs-nodes}
head(coefs(gimme_fit))
nodes(gimme_fit)
```

`coefs()` supplies the person-specific estimates that the prevalence
display abstracts away: one row per subject, layer, and path, holding that
subject's regression coefficient. The first rows belong to Aisha — her five
autoregressive coefficients, small values between −0.034 and 0.143, and the
first of her contemporaneous paths, efficacy predicting value at 0.48. The
`nodes()` table sums prevalence over the edges incident to each node within
a layer. Contemporaneously, efficacy is the most connected node (strength
2.125) and predominantly a sender (out-strength 1.875), while monitoring
and effort are the main receivers (in-strength 1.375 each) and value
participates least (strength 1.25). In the temporal layer the `self` column
records the autoregressive prevalence of 1 for every variable, with no
cross-variable strength.

```{r matrices}
matrices(gimme_fit)
```

`matrices()` returns the count and group-average coefficient matrices
behind these tables, with outcomes on the rows and predictors on the
columns. The temporal count matrix has 8 on its diagonal and 0 elsewhere;
the contemporaneous count matrix reaches 8 only in the two group cells. The
average coefficient matrices put magnitudes on the shared structure: across
the eight students the efficacy-to-monitoring path averages 0.325 and the
planning-to-effort path 0.361, while the mean autoregressive coefficients
lie between −0.027 and 0.045. A path can therefore be universal and weak —
at these cutoffs the shared temporal structure is exactly that — so
prevalence and magnitude have to be read together, and neither should be
mistaken for a standardized effect size.

# Visualizing the network

Plotting the fit draws the mixed network in the convention of the `gimme`
package: a single panel over the five nodes in which dashed edges are
lag-one temporal paths, solid edges are contemporaneous paths, black edges
belong to the group model, grey edges are individual-level, and edge width
scales with the weight. Under the default `weight = "prop"` the width is
path prevalence.

```{r plot-prop, eval=has_cograph}
plot(gimme_fit, weight = "prop")
```

The dashed black self-loops and the two solid black arrows — efficacy to
monitoring and planning to effort — constitute the group model; the grey
arrows are individual-level paths, drawn thinner in proportion to how few
students carry them.

```{r plot-coef, eval=has_cograph}
plot(gimme_fit, weight = "coef")
```

Reweighting by `weight = "coef"` keeps the same graph but scales width by
the group-average coefficient, so the display answers how strong rather
than how common: the two group contemporaneous arrows dominate at 0.325 and
0.361, and the autoregressive loops recede because their average
coefficients fall below 0.05 in absolute value.

# References
