---
title: "5. Multilevel VAR (mlVAR)"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{5. Multilevel VAR (mlVAR)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 5.5
)
```

```{r libraries}
library(idiographic)
data(srl)
vars <- c("efficacy", "value", "planning", "monitoring", "effort")
has_cograph <- requireNamespace("cograph", quietly = TRUE)
```

`build_mlvar()` estimates **group-level** temporal, contemporaneous, and
between-person networks in one multilevel model. Use it when the question moves
from one person's dynamics to the *average within-person process* and the
*between-person* differences across the whole sample.

# Fit

```{r mlvar}
mlvar_fit <- build_mlvar(srl, vars = vars, id = "name", standardize = TRUE)

mlvar_fit
```

Three networks are estimated and shown: the directed temporal network (average
within-person lag-1 effects), the undirected contemporaneous network, and the
undirected between-person network.

# Tidy tables

`edges()` stacks all three networks into one tidy table with a `network`
column; filter or summarise it with ordinary verbs.

```{r mlvar-tables}
head(edges(mlvar_fit))

summary(mlvar_fit)
```

```{r mlvar-matrices}
matrices(mlvar_fit)
```

# Plot

Draw the whole result, or any single layer by name — including the
between-person network.

```{r plot-mlvar, eval=has_cograph}
plot(mlvar_fit, layer = "temporal")
```

```{r plot-mlvar-between, eval=has_cograph}
plot(mlvar_fit, layer = "between")
```
