---
title: "Pooling Person-Specific Coefficients with pool_coefs()"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Pooling Person-Specific Coefficients with pool_coefs()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE, fig.width = 7.2, fig.height = 4.6, dpi = 144, fig.align = "center", out.width = "100%")
source("figure-style.R")
library(idiographic)
data(srl)
analysis_data <- subset(srl, name %in% c("Aisha", "Alice", "Anika", "Astrid", "Bjorn", "Bob", "Charlie", "Diana", "Erik", "Eve", "Fatima", "Frank"))
```

## Analytic question

`pool_coefs()` estimates the distribution of person-specific coefficients
after accounting for their sampling uncertainty. The function takes an
`idiographic_fit` with individual coefficients and standard errors. It returns
one tidy row per term with a pooled estimate, confidence interval, observed
coefficient spread, estimated between-person spread, Cochran's Q, and I-squared.

`pool_coefs()` applies a DerSimonian-Laird random-effects model
[@dersimonian1986]. Each learner's coefficient contributes according to its
sampling variance. The estimated `tau` describes between-person standard
deviation after sampling error is removed. The observed standard deviation
describes the raw spread before that correction.

## Fit individual regressions

`fit_lm()` supplies the person-specific coefficients and standard errors that
pooling requires. The individual scope fits effort on efficacy and monitoring
for each learner.

```{r pooling-fit}
individual_fit <- fit_lm(analysis_data, y = "effort", x = c("efficacy", "monitoring"), id = "name", time = "day", scope = "individual")
coefs(individual_fit, scope = "individual", n = 8)
```

`coefs()` shows that each learner contributes a separate estimate and standard
error in Table 1. Machine-learning coefficients do not generally carry
comparable standard errors, so `pool_coefs()` accepts LM, GLM, and
within-between results and rejects unsupported coefficient tables.

## Pool the coefficients

`pool_coefs()` estimates one random-effects summary for each regression term.
Table 2 reports all terms.

```{r pooling-table}
pooled_coefs <- pool_coefs(individual_fit)
pooled_coefs
```

`pool_coefs()` estimates an efficacy coefficient of about 0.29 in Table 2.
Its confidence interval excludes zero. The estimated between-person standard
deviation is about 0.20, and I-squared is about 0.81. The monitoring interval
includes zero and its I-squared is about 0.94. The panel therefore contains
substantial coefficient heterogeneity even where the pooled mean is uncertain.

## Plot the pooled coefficients

The forest plot displays the two substantive pooled slopes and their
confidence intervals. The intercept is omitted because its much larger scale
would compress the predictor effects. Heterogeneity remains available in
Table 2 through `tau` and I-squared.

```{r pooling-plot, fig.cap="Figure 1. Random-effects pooled slopes with 95% confidence intervals."}
shown_pool <- subset(pooled_coefs, term != "(Intercept)")
forest_plot(shown_pool$estimate, shown_pool$conf_low, shown_pool$conf_high,
            shown_pool$term, xlab = "Random-effects pooled coefficient (95% CI)")
```

Figure 1 shows a positive pooled efficacy slope and a monitoring interval that
crosses zero. These are average slopes, not claims of a common effect. The
large I-squared values in Table 2 show why the pooled estimates should be
reported together with their between-person heterogeneity.

## Assumptions and failure checks

`pool_coefs()` treats learners as independent units and assumes that their
true coefficients follow a distribution summarized by a mean and variance.
The DerSimonian-Laird estimate can be imprecise with few people. Q and
I-squared should not be read without the number of contributing learners and
the estimated `tau`.

`pool_coefs()` requires at least two usable person estimates for a term. A
coefficient with a missing or nonpositive standard error is excluded. Pooling
does not turn an associational coefficient into a causal effect.

## When to use which

`pool_coefs()` is appropriate for the average and heterogeneity of
person-specific slopes. `shrink_coefs()` is appropriate when the objective is
to stabilize each learner's estimate. `fit_lm(scope = "pooled")` estimates one
row-level pooled regression and answers a different question from
meta-analytic pooling of individual slopes.

## References
