---
title: "Panel Preparation with preprocess_panel()"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Panel Preparation with preprocess_panel()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE, fig.width = 7.2, fig.height = 4.6, dpi = 144, fig.align = "center", out.width = "100%")
source("figure-style.R")
library(idiographic)
data(srl)
analysis_data <- subset(srl, name %in% c("Aisha", "Alice", "Anika", "Astrid", "Bjorn", "Bob", "Charlie", "Diana", "Erik", "Eve", "Fatima", "Frank"))
```

## Analytic question

`preprocess_panel()` prepares repeated measurements without crossing person
boundaries. The function takes a long-format panel, person and time columns,
and variables to transform. It returns the original data frame with requested
centred, scaled, detrended, decomposed, or lagged columns.

`preprocess_panel()` treats centring and decomposition as different analytic
operations. Person centring replaces a variable with deviations from each
person's mean. Decomposition retains both the deviation and the person mean.
The latter supports separate within-person and between-person coefficients in
one pooled model [@mundlak1978].

## Create decomposition and lag columns

`preprocess_panel()` orders rows by learner and day before constructing lags.
The call below adds within-person and between-person components plus a
one-occasion lag for efficacy and monitoring. Table 1 displays the returned
columns directly.

```{r preparation-table}
prepared <- preprocess_panel(analysis_data, id = "name", time = "day", vars = c("efficacy", "monitoring"), decompose = TRUE, lag = 1)
head(prepared)
```

`preprocess_panel()` sets the first lag of each learner to missing in Table 1.
The second row for Aisha receives the first row's efficacy and monitoring
values. The within-person efficacy column expresses each daily value as a
deviation from Aisha's mean. The between-person column repeats that mean across
Aisha's rows.

## Plot the transformed series

The time-series display uses the centred values returned by
`preprocess_panel()`. A zero reference identifies Aisha's own mean, so the
sign of each point has a direct within-person interpretation.

```{r preparation-plot, fig.cap="Figure 1. Person-centred efficacy over time for Aisha."}
aisha <- subset(prepared, name == "Aisha")
figure_begin()
plot(aisha$day, aisha$efficacy_within, type = "l", lwd = 1.8,
     col = fig_col["blue"], xlab = "Day",
     ylab = "Deviation from Aisha's mean")
figure_grid(x = FALSE, y = TRUE)
abline(h = 0, col = fig_col["orange"], lwd = 1.4, lty = 2)
points(aisha$day, aisha$efficacy_within, pch = 21,
       bg = fig_col["blue"], col = "white", cex = 0.62)
```

Figure 1 fluctuates around zero because the series is person-centred. Positive
values mark days above Aisha's typical efficacy. Negative values mark days
below her typical efficacy. The transformation changes the interpretation of
a regression coefficient from a raw-score contrast to a within-person
contrast.

## Assumptions and failure checks

`preprocess_panel()` requires a meaningful ordering variable for lagging and
detrending. A lag represents the preceding recorded occasion. The
`lag_max_gap` argument can invalidate lags that span an unacceptable elapsed
time. No lag is carried from the final row of one learner to the first row of
another learner.

`preprocess_panel()` applies scaling parameters within the requested scope.
Person scaling changes coefficients into person-specific standard-deviation
units. Grand scaling uses a common scale. Detrending removes an estimated
linear trend and requires a time column. Transformation choices should follow
the estimand rather than a desire to improve model fit.

## When to use which

`preprocess_panel()` is appropriate for explicit centring, scaling,
decomposition, detrending, and lag construction. `preprocess()` is the
network-readiness audit for stationarity, compliance, and series variance.
`fit_within_between()` performs decomposition inside the model and is preferred
when the main question directly contrasts within-person and between-person
effects.

## References
