---
title: "Stabilizing Person-Specific Coefficients with shrink_coefs()"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Stabilizing Person-Specific Coefficients with shrink_coefs()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE, fig.width = 7.2, fig.height = 4.6, dpi = 144, fig.align = "center", out.width = "100%")
source("figure-style.R")
library(idiographic)
data(srl)
analysis_data <- subset(srl, name %in% c("Aisha", "Alice", "Anika", "Astrid", "Bjorn", "Bob", "Charlie", "Diana", "Erik", "Eve", "Fatima", "Frank"))
```

## Analytic question

`shrink_coefs()` stabilizes each person's coefficient using the pooled
coefficient distribution. The function takes a fit with individual estimates
and standard errors. It returns one tidy row per person and term with the raw
estimate, standard error, shrinkage weight, and shrunken estimate.

`shrink_coefs()` uses an empirical-Bayes rule. The shrinkage weight is the
estimated between-person variance divided by that variance plus the person's
sampling variance. A weight near one retains most of the person's raw
estimate. A weight near zero moves the estimate toward the pooled mean.

## Estimate individual slopes

`fit_lm()` estimates separate effort regressions for the 12 learners. The
efficacy term is retained because its person-specific slopes and standard
errors are the quantities of interest.

```{r shrinkage-fit}
individual_fit <- fit_lm(analysis_data, y = "effort", x = c("efficacy", "monitoring"), id = "name", time = "day", scope = "individual")
coefs(individual_fit, scope = "individual", n = 8)
```

`coefs()` displays raw person-specific estimates in Table 1. Differences among
these values contain both process heterogeneity and sampling error. Ranking
learners by the raw values would treat both sources as substantive.

## Apply empirical-Bayes shrinkage

`shrink_coefs()` estimates the pooled distribution and applies a learner-specific
weight. Table 2 retains only the efficacy term through a public argument.

```{r shrinkage-table}
shrunken_coefs <- shrink_coefs(individual_fit, term = "efficacy")
shrunken_coefs
```

`shrink_coefs()` moves every efficacy slope toward the pooled efficacy effect
in Table 2. Learners with larger standard errors receive smaller weights and
move farther. The shrunken values preserve estimated heterogeneity while
reducing the influence of measurement noise.

## Plot raw and shrunken estimates

The paired-dot display places learners on rows and coefficients on the
horizontal axis. Each line begins at the raw estimate and ends at the shrunken
estimate, making both the direction and size of shrinkage visible.

```{r shrinkage-plot, fig.cap="Figure 1. Raw and empirical-Bayes efficacy coefficients for 12 learners."}
shrunken_coefs <- shrunken_coefs[order(shrunken_coefs$shrunken), ]
at <- seq_len(nrow(shrunken_coefs))
limits <- range(c(shrunken_coefs$estimate, shrunken_coefs$shrunken, 0))
figure_begin(c(4.2, 7, 1, 1))
plot(shrunken_coefs$shrunken, at, xlim = limits,
     ylim = c(0.5, length(at) + 0.5), yaxt = "n", pch = 21,
     bg = fig_col["teal"], col = "white", cex = 1.3,
     xlab = "Efficacy coefficient", ylab = "")
figure_grid(x = TRUE, y = FALSE)
abline(v = 0, col = fig_col["muted"], lty = 2)
segments(shrunken_coefs$estimate, at, shrunken_coefs$shrunken, at,
         col = fig_col["grid"], lwd = 2.4)
points(shrunken_coefs$estimate, at, pch = 21, bg = "white",
       col = fig_col["muted"], cex = 1.05)
points(shrunken_coefs$shrunken, at, pch = 21, bg = fig_col["teal"],
       col = "white", cex = 1.3)
axis(2, at = at, labels = shrunken_coefs$subject, tick = FALSE,
     col.axis = fig_col["ink"])
legend("bottomright", c("Raw", "Shrunken"), pch = 21,
       pt.bg = c("white", fig_col["teal"]),
       col = c(fig_col["muted"], "white"), bty = "n", cex = 0.8)
```

Figure 1 shows the raw-to-shrunken movement explicitly. Extreme raw values
move towards the pooled centre, while precise estimates move less. The plot
should be interpreted with the weight column in Table 2: longer connecting
lines indicate that the individual's estimate carried less weight.

## Assumptions and failure checks

`shrink_coefs()` inherits the random-effects assumptions of `pool_coefs()`.
The method assumes independent people and a coefficient distribution that can
be summarized by one mean and variance. A multimodal coefficient distribution
can make one pooled centre inadequate.

`shrink_coefs()` requires standard errors. It does not accept coefficient
tables that assign equal certainty without evidence. Shrinkage estimates are
appropriate for description and stabilized prediction. Their usual intervals
require additional modelling if formal person-specific inference is the goal.

## When to use which

`shrink_coefs()` is appropriate for stabilized person-level estimates.
`pool_coefs()` is appropriate for the population mean and between-person
spread. `test_subgroups()` is appropriate before replacing one coefficient
distribution with discrete classes. Raw individual coefficients remain useful
when each person has enough information and minimal sampling error.

## References
