---
title: "Testing Whether Subgroups Exist with test_subgroups()"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Testing Whether Subgroups Exist with test_subgroups()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE, fig.width = 7.2, fig.height = 4.6, dpi = 144, fig.align = "center", out.width = "100%")
source("figure-style.R")
library(idiographic)
data(srl)
analysis_data <- subset(srl, name %in% c("Aisha", "Alice", "Anika", "Astrid", "Bjorn", "Bob", "Charlie", "Diana", "Erik", "Eve", "Fatima", "Frank"))
```

## Analytic question

`test_subgroups()` asks whether person-specific coefficient vectors are better
represented by one population or by several discrete groups. The function
takes a repeated-measures regression problem and a maximum group count. It
returns a selected count, an evidence label, distribution-shape diagnostics,
and a tidy BIC table.

`test_subgroups()` fits person-specific regressions and compares Gaussian
mixtures with one through `k_max` components. The search covers several
covariance structures. Bayesian information criterion balances mixture fit
against model complexity. A clustering algorithm always returns a partition
when asked for one. This test allows one population to remain the selected
answer.

## Test one through three groups

`test_subgroups()` assesses efficacy and monitoring slopes across 12 learners.
The printed object states the selected group count and evidence relative to one
group.

```{r subgroup-test}
subgroup_test <- test_subgroups(analysis_data, y = "effort", x = c("efficacy", "monitoring"), id = "name", time = "day", k_max = 3)
subgroup_test
```

`test_subgroups()` selects three groups for this 12-learner subset. The printed
verdict reports a BIC improvement of about 8.9 over the one-group model. This
result supports further inspection of a three-group representation. It does
not assign learners to groups.

## Inspect the model-selection table

`as.data.frame()` returns the BIC search as one row per fitted mixture. Table 1
marks the selected row.

```{r subgroup-selection-table}
as.data.frame(subgroup_test)
```

`as.data.frame()` reports a minimum BIC of about 61.1 for the selected
three-component model in Table 1. The best one-component model has a BIC of
about 70.1. Model selection depends on the coefficient distribution and the
number of learners.

## Plot the BIC comparison

The figure shows every fitted covariance model faintly and connects the best
BIC at each candidate group count. This separates the group-count decision
from the nuisance choice of covariance structure.

```{r subgroup-test-plot, fig.cap="Figure 1. Bayesian information criterion across candidate subgroup models."}
bic_table <- as.data.frame(subgroup_test)
best_by_k <- aggregate(bic ~ k, data = bic_table, min)
figure_begin()
plot(bic_table$k, bic_table$bic, pch = 21, bg = fig_col["grid"],
     col = "white", cex = 1.0, xaxt = "n",
     xlab = "Candidate number of subgroups", ylab = "BIC")
figure_grid(x = FALSE, y = TRUE)
lines(best_by_k$k, best_by_k$bic, col = fig_col["blue"], lwd = 2)
points(best_by_k$k, best_by_k$bic, pch = 21, bg = fig_col["blue"],
       col = "white", cex = 1.35)
selected <- subset(bic_table, selected)
points(selected$k, selected$bic, pch = 21, bg = fig_col["orange"],
       col = "white", cex = 1.65)
text(selected$k, selected$bic, "selected", pos = 3, cex = 0.75,
     col = fig_col["orange"])
axis(1, at = sort(unique(bic_table$k)))
```

Figure 1 places the selected three-group solution at the global BIC minimum.
The faint points show that covariance assumptions materially affect the score;
the connected blue series shows the best available evidence for each group
count. Lower BIC indicates the preferred fit-complexity trade-off.

## Assumptions and failure checks

`test_subgroups()` assumes that the person-specific coefficient distribution
can be compared with Gaussian mixtures. Skewness and heavy tails can produce
apparent classes when the population is continuous. Bauer and Curran describe
this overextraction problem for mixture models [@bauer2003]. The returned shape
diagnostics should therefore accompany the BIC result.

`test_subgroups()` needs more people than candidate components and enough
observations to estimate each person's slopes. A selected class count is
sample-dependent. Replication or stability analysis is needed before treating
classes as substantive categories.

## When to use which

`test_subgroups()` is appropriate for evidence about the number of coefficient
populations. `find_subgroups()` is appropriate for assigning people to a
requested or selected number of groups. `fit_subgroups()` is appropriate for
comparing group-specific prediction with pooled and individual alternatives.

## References
