---
title: "From Raw Panel Data to Person-Specific Conclusions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From Raw Panel Data to Person-Specific Conclusions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(idiographic)
```

`idiographic` treats the person as the primary analytical unit. This vignette
shows the common path from inspecting repeated observations to comparing
person-specific conclusions. Network methods use the same person-by-time data
contract and are covered in the method-specific vignettes.

## Inspect the panel

The bundled `srl` data contain daily observations nested within learners.

```{r inspect}
vars <- c("efficacy", "monitoring", "effort")

describe_persons(srl, id = "name", vars = vars, time = "day", n = 6)
correlate_persons(srl, id = "name", vars = vars, n = 6)
variance_components(srl, id = "name", vars = vars)
```

The descriptive layer reports each person's usable observations, missingness,
variation, serial dependence, and sampling gaps. Inspecting these quantities
before fitting helps distinguish a model failure from a panel that cannot
identify the requested person-specific effect.

## Prepare within-person variables

`preprocess_panel()` transforms columns without crossing person boundaries.
The example adds a person-centred predictor and its previous-occasion value.

```{r prepare}
panel <- preprocess_panel(
  srl,
  id = "name",
  time = "day",
  vars = c("efficacy", "monitoring"),
  center = "person",
  lag = 1
)
```

Use `preprocess()` instead when the goal is the established network-readiness
audit (stationarity, compliance, variance, and related diagnostics).

## Compare pooled and individual estimates

The default scope fits a pooled comparison and one model per person. Ordered
hold-out validation uses the final observations of each person as test rows.

```{r regression}
fit <- fit_lm(
  panel,
  y = "effort",
  x = c("efficacy", "monitoring", "efficacy_lag1"),
  id = "name",
  time = "day",
  scope = "both",
  min_train = 30
)

metrics(fit, overall = TRUE)
coefs(fit, scope = "individual", n = 8)
```

`individuals(fit)`, `pooled(fit)`, `person(fit, id)`, and `overall(fit)` create
focused views without changing the underlying estimates.

## Separate within-person and between-person effects

Raw panel regressions can mix two different questions: whether people with a
higher typical predictor value have a higher outcome, and whether a person has
a higher outcome than usual when their predictor is higher than usual.

```{r within-between}
wb <- fit_within_between(
  srl,
  y = "effort",
  x = c("efficacy", "monitoring"),
  id = "name",
  time = "day",
  scope = "pooled",
  min_train = 30
)

contextual(wb)
```

The contextual table keeps these components explicit rather than silently
assigning one interpretation to a raw-score coefficient.

## Stabilise and explain differences between people

Individual coefficients vary because processes differ and because estimates
are noisy. Pooling estimates the coefficient distribution; shrinkage produces
stabilised person-specific coefficients.

```{r pooling}
pooled_coefs <- pool_coefs(fit)
shrunk_coefs <- shrink_coefs(fit)

pooled_coefs
head(shrunk_coefs)
```

Subgroup and treatment-effect workflows follow the same principle: test what
the data identify, retain person-level uncertainty and failures, and use pooled
quantities as comparisons rather than replacements for individual processes.
See `test_subgroups()`, `find_subgroups()`, `fit_effects()`, and
`fit_heterogeneity()` for those analyses.

## Move to dynamic networks when the question is multivariate

When the target is a system of lagged and contemporaneous relations rather than
one outcome, use the network estimators on the same panel:

```{r network, eval=FALSE}
net <- fit_mlvar(
  srl,
  vars = c("efficacy", "monitoring", "effort"),
  id = "name",
  beep = "day"
)

edges(net)
coefs(net)
```

The regression and network layers deliberately share `coefs()` and a common
idiographic vocabulary while retaining result structures appropriate to their
different estimands.
