---
title: "Within- and Between-Person Variance with variance_components()"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Within- and Between-Person Variance with variance_components()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE, fig.width = 7.2, fig.height = 4.6, dpi = 144, fig.align = "center", out.width = "100%")
source("figure-style.R")
library(idiographic)
data(srl)
analysis_data <- subset(srl, name %in% c("Aisha", "Alice", "Anika", "Astrid", "Bjorn", "Bob", "Charlie", "Diana", "Erik", "Eve", "Fatima", "Frank"))
```

## Analytic question

`variance_components()` quantifies whether variation occurs mainly between
people or within a person over time. The function takes a repeated-measures
data frame, one or more variables, and a person identifier. It returns the
within-person variance, between-person variance, total variance,
intraclass correlation, and reliability of the person mean.

`variance_components()` defines the intraclass correlation as the share of
total variance attributable to stable differences between people. A variable
with an intraclass correlation near zero varies mainly within people. A
variable with an intraclass correlation near one differs mainly between
people.

## Decompose the observed variance

`variance_components()` uses the analysis-of-variance estimator by default.
The optional REML estimator requires `lme4`. Table 1 reports the default
decomposition for three self-regulated-learning variables.

```{r variance-table}
components <- variance_components(analysis_data, vars = c("efficacy", "monitoring", "effort"), id = "name")
components
```

`variance_components()` assigns intraclass correlations of approximately
0.40 to efficacy, 0.38 to monitoring, and 0.26 to effort in Table 1. Most
observed variation therefore occurs within learners for all three variables.
Stable between-learner differences remain large enough to make a raw pooled
slope potentially mix levels of association.

## Plot the variance shares

`plot_variance()` converts each intraclass correlation into complementary
between-person and within-person shares. Figure 1 uses blue for between-person
variance and orange for within-person variance.

```{r variance-plot, fig.cap="Figure 1. Within-person and between-person shares of observed variance."}
plot_variance(components)
```

Figure 1 shows that the within-person share exceeds the between-person share
for each variable. Effort has the largest within-person share among the three.
This pattern supports models of daily fluctuation while retaining a separate
term for stable person means.

## Assumptions and failure checks

`variance_components()` assumes that person identifiers define independent
clusters. The default decomposition uses observed group means and pooled
within-group variation. The reliability column depends on both the
intraclass correlation and the number of observations per person. High
reliability of a person mean does not imply high within-person reliability of
individual occasions.

`variance_components()` requires variation across people to estimate a
between-person component. A variable that is constant within each person can
support a between-person comparison but cannot support a person-specific
time-varying slope. A variable with no between-person variation cannot identify
a contextual contrast.

## When to use which

`variance_components()` is appropriate before choosing a level of analysis.
`describe_persons()` is appropriate for person-specific data quality and
temporal movement. `fit_within_between()` is appropriate when the variance
decomposition motivates separate within-person and between-person
coefficients. `pool_coefs()` is appropriate after estimating individual slopes
and addresses heterogeneity of effects rather than heterogeneity of variable
levels.

## References
